\bf Mechanistic models for panel data: analysis of ecological experiments with four interacting species

### 2024-04-17 ms.Rnw
Update
Update(unfinished)

\bigskip
\begin{abstract}
In ecological research, panel data, often comprising multi-magnitude time series, presents unique analytical challenges due to its high dimensionality. This study employs the PanelPOMP framework and the Panel Iterated Filtering (PIF) method to explore the practicality and advancement of mechanistic model development for such complex datasets. Mechanistic models, especially valuable in ecological research, allow for quantitative analysis and deeper understanding of dynamic systems. However, they are often constrained by the non-Gaussian nature of ecological data. Our study applies these methods to the \textit{Daphnia} panel data, consisting of independent time series with 10 data points each, to derive parameters that minimize the Akaike Information Criterion (AIC). The comparison of AIC values across models reveals the significant impact of alga, the prey of \textit{Daphnia}, and parasite on their population dynamics which is observed in the SIRPF model highlights this relationship. The panelPOMP framework's ability to handle complex panel data analysis is demonstrated, emphasizing its utility in uncovering latent processes not captured by traditional models. However, the study also discusses limitations and interpretations of mechanistic models and PanelPOMP framework in ecological data analysis.
\end{abstract}

\noindent%
{\it Keywords:}  Panel Iterated Filtering; SIRPF model; PanelPOMP; Daphnia.
\vfill

@@

Ecological experiments contain many elements or processes that are difficult to observe, models that can describe both observable and unobservable processes to explain the whole ecological experiment are necessary. As a mathematical description of the elements forming a system, mechanistic modeling can describe the mutual interactions between elements in the system and their interactions with the environment \citep{Stalidzans2020}. Utilizing the established process, they elucidate the actions of hidden mechanisms \citep{Duarte2003}. The purpose of mechanistic modeling in ecology involves employing mathematical models to depict the development and interaction of dynamic systems. This quantitative tool allows researchers to evaluate the explanatory potential of various conceptual models and measure the magnitude of pertinent parameters. Mechanistic modeling has been widely employed in social sciences, industrial engineering, and physics for an extended period. Improved computational capabilities now make it more effective. Traditional time series analysis offers limited explanatory power due to the nonstationary, nonlinear, and stochastic nature of ecological information. Despite the creation of theoretical models for the data, connecting theories with data continues to be a challenging task. As a result, ecological mechanical modeling cannot solve most of the problems on a wide scale, but rather, only a limited range of the problems \citep{levin1992problem}.\\

>>>>

\bigskip
\begin{abstract}
In ecological research, panel data usually consists of multiple time series, which brings challenges for analysis due to their high dimensionality. This study applies the PanelPOMP framework and the Panel Iterated Filtering (PIF) method to explore the utility and complexity of mechanistic models in complex system. Mechanistic models are particularly important in ecological research for quantitative analysis and deeper understanding of dynamic systems. However, they are often limited by the non-Gaussian nature of ecological data. Our study applies these methods to \textit{Daphnia} panel data (consisting of independent time series with 10 data points each) to derive parameters that minimize the Akaike information criterion (AIC). By utilizing likelihood ratio test and comparing the AIC values of the different models, a significant effect of algae (prey of \textit{Daphnia}) and parasites on their population dynamics was observed, while the SIRPF model highlighted this relationship. The study demonstrates the ability of the panelPOMP framework to handle complex panel data analyses, emphasizing its utility in revealing underlying processes not captured by traditional models. However, this study also discusses the limitations and interpretations of mechanistic models and the panelPOMP framework in ecological data analysis.
\end{abstract}

\noindent%
{\it Keywords:  Panel Iterated Filtering; SIRPF model; PanelPOMP; Daphnia}
\vfill

@@

Ecological experiments contain many elements or processes that are difficult to observe, so models that can describe both observable and unobservable processes are needed to explain the entire ecological experiment. As mathematical descriptions of the elements that make up a system, mechanistic models can describe the interactions between the elements of the system and their interactions with the environment. Using established processes, they can elucidate the role of hidden mechanisms \citep{Duarte2003}. The purpose of mechanistic modeling in ecology is to use mathematical models to describe the development and interactions of dynamic systems. This quantitative tool allows researchers to assess the explanatory potential of various conceptual models and to measure the magnitude of relevant parameters. Mechanistic modeling has been widely used in the social sciences, industrial engineering and physics. Now, increased computational power has made it even more effective. Because of the non-smooth, non-linear and stochastic nature of ecological information, traditional time-series analysis offers limited explanatory power. Despite creating theoretical models for the data, linking the theory to the data remains a challenging task. As a result, eco-mechanical modeling is unable to address most problems on a large scale, but only a limited range of problems \citep{levin1992problem}.\\
